---
title: "Lab 9"
author: "Joshua Eisner"
output: pdf_document
---

#YARF

For the next couple of labs, I want you to make some use of a package I wrote that offers convenient and flexible tree-building and random forest-building. Make sure you have a JDK installed first

https://www.oracle.com/java/technologies/downloads/

Then try to install rJava

```{r}
options(java.parameters = "-Xmx8000m")
pacman::p_load(rJava)
.jinit()
```

If you have error, messages, try to google them. Everyone has trouble with rJava!

If that worked, please try to run the following which will install YARF from my github:

```{r}
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
pacman::p_load(YARF)
```

Please try to fix the error messages (if they exist) as best as you can. I can help on slack.


# Missing Data

Load up the Boston Housing Data and separate into matrix `X` for the features and vector `y` for the response. Randomize the rows

```{r}
rm(list = ls())
set.seed(1)
boston = MASS::Boston
boston_shuffled = boston[sample(1 : nrow(boston)), ]
X = as.matrix(boston_shuffled[, 1 : 13])
y = boston_shuffled$medv
rm(boston, boston_shuffled)
```



Similar to lab 1, write a function that takes a matrix and punches holes (i.e. sets entries equal to `NA`) randomly with an argument `prob_missing`.

```{r}
punch_holes = function(mat, prob_missing){
  n = nrow(mat) * ncol(mat)
  is_missing = as.logical(rbinom(n, 1, prob_missing))
  mat[is_missing] = NA
  mat
}
```

Create a matrix `Xmiss` which is `X` but has missingness with probability of 10% using the function you just wrote. 

```{r}
Xmiss = punch_holes(X, .1)
```

What type of missing data mechanism created the missingness in `Xmiss`?

Imping

Also, generate the M matrix and delete columns that have no missingness.

```{r}
M = apply(is.na(Xmiss), 2, as.numeric)
colnames(M) = paste("is_missing_", colnames(X), sep = "")
M = M[, colSums(M) > 0]
```

Split the first 400 observations were the training data and the remaining observations are the test set. For Xmiss, cbind on the M so the model has a chance to fit on "is missing" as we discussed in class.

```{r}
train_idx = 1 : 400
test_idx = setdiff(1 : nrow(X), train_idx)
X_train =     X[train_idx, ]
Xmiss_train = cbind(Xmiss, M)[train_idx, ]
y_train =     y[train_idx]
X_test =      X[test_idx, ]
Xmiss_test =  cbind(Xmiss, M)[test_idx, ]
y_test =      y[test_idx]
```

Fit a random forest model of `y_train ~ X_train`, report oos s_e (not oob) on `X_test`. This ignores missingness

```{r}
model_rf = YARF(data.frame(X_train), y_train)
y_hat_test = predict(model_rf, data.frame(X_test))
sqrt(mean((y_hat_test - y_test)^2))
```

Impute the missingness in `Xmiss` using the feature averages to create a matrix `Ximp_naive_train` and `Ximp_naive_test`. 

```{r}
Ximp_naive_train = Xmiss_train
Ximp_naive_test = Xmiss_test
x_averages = array(NA, ncol(X))
for( j in 1: ncol(X)){
  x_averages[j] = mean(Xmiss_train, na.rm = TRUE)
  Ximp_naive_train[is.na(Xmiss_train[, j]), j] = x_averages[j]
  Ximp_naive_test[is.na(Xmiss_test[, j]), j] = x_averages[j]
}
```

Fit a random forest model of `y_train ~ Ximp_naive_train`, report oos s_e (not oob) on `Ximp_naive_test`.

```{r}
model_rf = YARF(data.frame(Ximp_naive_train), y_train)
y_hat_test = predict(model_rf, data.frame(Ximp_naive_test))
sqrt(mean((y_hat_test - y_test)^2))
```

How much predictive performance was lost due to missingness when naive imputation was used vs when there was no missingness?

```{r}

```

Use `missForest` to impute the missing entries to create a matrix `Ximp_MF_train` and `Ximp_MF_test`.

```{r}
pacman::p_load(missForest)
Xymiss = rbind(
  cbind(Xmiss_train, y_train),
  cbind(Xmiss_test, NA)
)
Xyimp_miss = missForest(Xymiss)$ximp
Ximp_MF_train = Xyimp_miss[train_idx, 1 : ncol(X)]
Ximp_MF_test = Xyimp_miss[test_idx, 1 : ncol(X)]
```

Fit a random forest model of `y_train ~ Ximp_MF_train`, report oos s_e (not oob) on `Ximp_MF_test`.

```{r}
model_rf = YARF(data.frame(Ximp_MF_train), y_train)
y_hat_test = predict(model_rf, data.frame(Ximp_MF_test))
sqrt(mean((y_hat_test - y_test)^2))
```

How much predictive performance was lost due to missingness when `missForest` imputation was used?

```{r}
.2
```

Why did `missForest` imputation perform better than naive imputation?

Bc it's in sample

Reload the feature matrix:

```{r}
rm(list = ls())
X = as.matrix(MASS::Boston[, 1 : 13])
```

Create missingness in the feature `lstat` that is due to a MAR missing data mechanism.

```{r}
#TO-DO
```

Create missingness in the feature `rm` that is a NMAR missing data mechanism.

```{r}
#TO-DO
```


#Bagged Trees and Random Forest

Take a training sample of n = 2000 observations from the diamonds data.

```{r}
rm(list = ls())
pacman::p_load(tidyverse)
set.seed(1)
diamonds_train = ggplot2::diamonds %>% 
  sample_n(2000)
y_train = diamonds_train$price
X_train = diamonds_train %>% dplyr::select(-price)
```


Using the diamonds data, find the oob s_e for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can create the bagged tree model via setting an argument within the RF constructor function. Plot.

```{r}
num_trees_values = c(1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400)
oob_se_bagged_trees_mod_by_num_trees = array(NA, length(num_trees_values))
for (m in 1 : length(num_trees_values)){
  mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees_values[m])
  oob_se_bagged_trees_mod_by_num_trees[m] = mod_bag$rmse_oob
}

ggplot(data.frame(num_trees = num_trees_values, oob_se = oob_se_bagged_trees_mod_by_num_trees)) +
  aes(x = num_trees, y = oob_se) +
  geom_point() + geom_line() + scale_x_log10()
```

Find the bootstrap s_e for a RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can calculate oob residuals via `e_oob = y_train - rf_mod$predicted`. Plot.

```{r}
oob_se_rf_mod_by_num_trees = array(NA, length(num_trees_values))
for (m in 1 : length(num_trees_values)){
  mod_rf = YARF(X_train, y_train, num_trees = num_trees_values[m])
  oob_se_rf_mod_by_num_trees[m] = mod_rf$rmse_oob
}


ggplot(data.frame(num_trees = num_trees_values, oob_se = oob_se_rf_mod_by_num_trees)) +
  aes(x = num_trees, y = oob_se) +
  geom_point() + geom_line() + scale_x_log10()
```

What is the percentage gain / loss in performance of the RF model vs bagged trees model for each number of trees? Gains are negative (as in lower oos s_e).

```{r}
cbind(
  num_trees_values,
  (oob_se_rf_mod_by_num_trees - oob_se_bagged_trees_mod_by_num_trees) / oob_se_bagged_trees_mod_by_num_trees * 100
)
```

Why was this the result?

#TODO

Plot oob s_e by number of trees for both RF and bagged trees by creating a long data frame from the two results.

```{r}
wide_df = data.table(
  num_trees = num_trees_values,
  oob_se_bag = oob_se_bagged_trees_mod_by_num_trees,
  oob_se_rf = oob_se_rf_mod_by_num_trees
)
long_df = melt(wide_df, measure.vars = c("oob_se_bag", "oob_se_rf")) 

ggplot(long_df) + 
  geom_line(aes(x = num_trees, y = value, color = variable))
```

Build RF models for 500 trees using different `mtry` values: 1, 2, ... the maximum. That maximum will be the number of features assuming that we do not binarize categorical features if you are using `randomForest` or the number of features assuming binarization of the categorical features if you are using `YARF`. Calculate oob s_e for all mtry values.

```{r}
oob_se_by_mtry = array(NA, ncol(diamonds_train))
#TO-DO
```

Plot oob s_e by mtry.

```{r}
#TO-DO
```

Take a sample of n = 2000 observations from the adult data and name it `adult_sample`. Then impute missing values using missForest.

```{r}
rm(list = ls())
pacman::p_load(missForest)
set.seed(1)
pacman::p_load_gh("coatless/ucidata")
adult_train = adult %>% 
  sample_n(2000)
adult_train = missForest(adult_train)$ximp
X_train = adult_train %>%
  select(-income)
y_train = adult_train$income
```


Using the adult_train data, find the bootstrap misclassification error for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. Plot.

```{r}
num_trees_values = c(1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000)
oob_se_bagged_trees_mod_by_num_trees = array(NA, length(num_trees_values))
for (m in 1 : length(num_trees_values)){
  mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees_values[m])
  oob_se_bagged_trees_mod_by_num_trees[m] = mod_bag$misclassification_error
}
ggplot(data.frame(num_trees = num_trees_values, oob_se = oob_se_bagged_trees_mod_by_num_trees)) +
  aes(x = num_trees, y = oob_se) +
  geom_point() + geom_line() + scale_x_log10()
```

Using the adult_train data, find the bootstrap misclassification error for an RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees.

```{r}
oob_se_rf_mod_by_num_trees = array(NA, length(num_trees_values))
#TO-DO
```

What is the percentage gain / loss in performance of the RF model vs bagged trees model?

```{r}
cbind(
  num_trees_values,
  (oob_se_rf_mod_by_num_trees - oob_se_bagged_trees_mod_by_num_trees) / oob_se_bagged_trees_mod_by_num_trees * 100
)
```

Build RF models on adult_train for 500 trees using different `mtry` values: 1, 2, ... the maximum (see above as maximum is defined by the specific RF algorithm implementation). 

```{r}
oob_se_by_mtry = array(NA, ncol(adult_train))
#TO-DO
```


Plot bootstrap misclassification error by `mtry`.

```{r}
#TO-DO
```

Is `mtry` an important hyperparameter to optimize when using the RF algorithm? Explain

#TO-DO

Identify the best model among all values of `mtry`. Fit this RF model. Then report the following oob error metrics: misclassification error, precision, recall, F1, FDR, FOR and compute a confusion matrix.

```{r}
#TO-DO
```

Is this a good model? (yes/no and explain).

#TO-DO

There are probability asymmetric costs to the two types of errors. Assign two costs below and calculate oob total cost.

```{r}
fp_cost = 
fn_cost = 
#TO-DO
```

# Asymmetric Cost Modeling, ROC and DET curves

Fit a logistic regression model to the adult_train missingness-imputed data.

```{r}
rm(list = setdiff(ls(), "adult_train"))
#TO-DO
```

Use the function from class to calculate all the error metrics (misclassification error, precision, recall, F1, FDR, FOR) for the values of the probability threshold being 0.001, 0.002, ..., 0.999 in a tibble (dplyr data frame).

```{r}
pacman::p_load(tidyverse)
asymmetric_predictions_results = tibble(
  p_hat_threshold = seq(from = 0.001, to = 0.999, by = 0.001),
  misclassification_error = NA, 
  precision = NA, 
  recall = NA, 
  F1 = NA, 
  FDR = NA, 
  FOR = NA
)
#TO-DO
```

Calculate the column `total_cost` and append it to this data frame via `mutate`.

```{r}
#TO-DO
```

Which is the lowest total cost? What is the "winning" probability threshold value providing that minimum total cost?

```{r}
#TO-DO
```

Plot an ROC curve and interpret.

```{r}
#TO-DO
```

#TO-DO interpretation

Calculate AUC and interpret.

```{r}
#TO-DO
```

#TO-DO interpretation

Plot a DET curve and interpret.

```{r}
#TO-DO
```

#TO-DO interpretation
